---
title: "Data Exploration"
author: "Joffrey JOUMAA"
date: "`r format(Sys.Date(), format = '%d %B %Y')`"
output:
  bookdown::html_document2:
    css: cosmo_custom.css
    number_sections: yes
    code_folding: show
    df_print: default
    fig_caption: yes
    code_download: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
vignette: >
  %\VignetteIndexEntry{Data Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
# command to build package without getting vignette error
# https://github.com/rstudio/renv/issues/833
# devtools::check(build_args=c("--no-build-vignettes"))

# reduce png size
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

# global option relative to rmarkdown
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  tidy = TRUE,
  optipng = "-o7 -quiet",
  pngquant = "--speed=1"
)

# library
library(data.table)
library(ggplot2)
library(kableExtra)
library(leaflet)
library(gtsummary) # https://cran.r-project.org/web/packages/gtsummary/vignettes/tbl_summary.html
library(corrplot)
library(ggcorrplot)
library(ggnewscale)
library(magrittr)
library(DT)

# remove some warnings
suppressWarnings(library(ggplot2))

# define my own table format: https://github.com/haozhu233/kableExtra/issues/374
sable <- function(x, escape = T, ...) {
  knitr::kable(x, escape = escape, ...) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "responsive"),
      full_width = F
    )
}

# theme ggplot
# based: https://benjaminlouis-stat.fr/en/blog/2020-05-21-astuces-ggplot-rmarkdown/
theme_jjo <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # the whole figure
      # plot.title = element_text(size = rel(1), face = "bold", margin = margin(0,0,5,0), hjust = 0),
      # figure area
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      # axes
      # axis.title = element_text(size = rel(0.85), face = "bold"),
      # axis.text = element_text(size = rel(0.70), face = "bold"),
      axis.line = element_line(color = "black", arrow = arrow(length = unit(0.2, "lines"), type = "closed")),
      # legend
      # legend.title = element_text(size = rel(0.85), face = "bold"),
      # legend.text = element_text(size = rel(0.70), face = "bold"),
      # legend.key = element_rect(fill = "transparent", colour = NA),
      # legend.key.size = unit(1.5, "lines"),
      # legend.background = element_rect(fill = "transparent", colour = NA),
      # Les <U+00E9>tiquettes dans le cas d'un facetting
      strip.background = element_rect(fill = "#888888", color = "#888888"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "white", margin = margin(5, 0, 5, 0))
    )
}
```

## {.unnumbered}

This document aims at exploring two datasets, one in 2016 on 6 individuals and another one in 2018 on 4 individuals. For that purpose, we need first to load the `weanlingNES` package to load data.

```{r}
# load library
library(weanlingNES)

# load data
data("data_nes", package = "weanlingNES")
# load("../data/data_nes.rda")
```

## 2016-individuals 

Let's have a look at what's inside `data_nes$data_2016`:
  
```{r}
# list structure
str(data_nes$year_2016, max.level = 1, give.attr = F, no.list = T)
```

> A list of `r length(data_nes$year_2016)` `data.frames`, one for each seal

For convenience, we aggregate all `r length(data_nes$year_2016)` individuals into one dataset.

```{r, eval=FALSE}
# combine all individuals
data_2016 <- rbindlist(data_nes$year_2016, use.name = TRUE, idcol = TRUE)

# display
DT::datatable(data_2016[sample.int(.N, 100), ], options = list(scrollX = T))
```
```{r, echo=FALSE, results='asis'}
# combine all individuals
data_2016 <- rbindlist(data_nes$year_2016, use.name = TRUE, idcol = TRUE)

# title
cat("<table style='width: 50%'>", paste0("<caption>", "(#tab:myDThtmltools)", "Sample of 100 random rows from `data_2016`", "</caption>"), "</table>", sep = "\n")

# display
DT::datatable(data_2016[sample.int(.N, 100), ], options = list(scrollX = T))
```

### Summary

```{r}
# raw_data
data_2016[, .(
  nb_days_recorded = uniqueN(as.Date(date)),
  max_depth = max(maxpress_dbars),
  sst_mean = mean(sst2_c),
  sst_sd = sd(sst2_c)
), by = .id] %>%
  sable(
    caption = "Summary diving information relative to each 2016 individual",
    digits = 2
  )
```

Well, it seems that `sst` is a bit odd. Let's have a look at its distribution.

```{r, fig.cap="Distribution of raw `sst2` for the four individuals in 2016"}
ggplot(data_2016, aes(x = sst2_c, fill = .id)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(.id ~ .) +
  theme_jjo()
```

Let's remove any data with a `sst2_c` > 500.

```{r, fig.cap="Distribution of filtered `sst2` for the four individuals in 2016"}
data_2016_filter <- data_2016[sst2_c < 500, ]
ggplot(data_2016_filter, aes(x = sst2_c, fill = .id)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(.id ~ .) +
  theme_jjo()
```

Well, that seems to be much better! In the process of filtering out odd values, we removed `r data_2016[,.N] - data_2016_filter[,.N]` rows this way:

```{r}
# nbrow removed
data_2016[sst2_c > 500, .(nb_row_removed = .N), by = .id] %>%
  sable(caption = "# of rows removed by 2016-individuals")
```
```{r, fig.cap="Where and when the `sst2` outliers occured", fig.width=9}
# max depth
ggplot(
  data_2016,
  aes(y = -maxpress_dbars, x = as.Date(date), col = .id)
) +
  geom_path(show.legend = FALSE) +
  geom_point(data = data_2016[sst2_c > 500, ], col = "black") +
  scale_x_date(date_labels = "%m/%Y") +
  labs(y = "Pressure (dbar)", x = "Date") +
  facet_wrap(.id ~ .) +
  theme_jjo() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Well this latter plot highlights several points:

* most of the outliers occur while animals spend the whole day at the surface (or on the ground), probably resting, so essentially at the beginning and the end of each track
* we can already see that `ind_3449` seems to have return ashore twice during his track

Let's see if we can double check that, using a map.

```{r, fig.cap="It is supposed to be the track of `ind_3449`... (red dots are the location of removed rows)", fig.width=8}
# interactive map
leaflet() %>%
  setView(lng = -122, lat = 38, zoom = 2) %>%
  addTiles() %>%
  addPolylines(
    lat = data_2016[.id == "ind_3449", latitude_degs],
    lng = data_2016[.id == "ind_3449", longitude_degs],
    weight = 2
  ) %>%
  addCircleMarkers(
    lat = data_2016[.id == "ind_3449" & sst2_c > 500, latitude_degs],
    lng = data_2016[.id == "ind_3449" & sst2_c > 500, longitude_degs],
    radius = 3,
    stroke = FALSE,
    color = "red",
    fillOpacity = 1
  )
```

... these coordinates seem weird !

```{r}
# summary of the coordinates by individuals
data_2016[, .(.id, longitude_degs, latitude_degs)] %>%
  tbl_summary(by = .id) %>%
  modify_caption("Summary of `longitude_degree` and `latitude_degree`")
```

```{r, fig.width=9, fig.cap="Distribution of coordinates per seal"}
# distribution coordinates
ggplot(
  data = melt(data_2016[, .(
    Longitude = longitude_degs,
    Latitude = latitude_degs,
    .id
  )],
  id.vars =
    ".id", value.name = "Coordinate"
  ),
  aes(x = Coordinate, fill = .id)
) +
  geom_histogram(show.legend = F) +
  facet_grid(variable ~ .id) +
  theme_jjo()
```

There is definitely something wrong with these coordinates (five seals would have crossed the equator...), but the representation of the track can also be improved! Here are the some points to explore:

* For `longitude` a part of the data seems to have a wrong sign, resulting in these distribution, that appear to be cut off
* For `latitude`, well this is ensure but maybe the same problem occurs

Let's try to play on coordinates' sign to see if we can display something that makes more sense.

```{r, fig.cap="An attempt to display the `ind_3449`'s track", fig.width=8}
# interactive map
leaflet() %>%
  setView(lng = -122, lat = 50, zoom = 3) %>%
  addTiles() %>%
  addPolylines(
    lat = data_2016[.id == "ind_3449", abs(latitude_degs)],
    lng = data_2016[.id == "ind_3449", -abs(longitude_degs)],
    weight = 2
  )
```

> I'll better ask Roxanne!
  
### Missing values
  
```{r fig.cap="Check for missing value in 2016-individuals", out.width="100%"}
# build dataset to check for missing values
dataPlot <- melt(data_2016_filter[, .(.id, is.na(.SD)), .SDcol = -c(
  ".id",
  "rec#",
  "date",
  "time"
)])
# add the id of rows
dataPlot[, id_row := c(1:.N), by = c("variable", ".id")]

# plot
ggplot(dataPlot, aes(x = variable, y = id_row, fill = value)) +
  geom_tile() +
  labs(x = "Attributes", y = "Rows") +
  scale_fill_manual(
    values = c("white", "black"),
    labels = c("Real", "Missing")
  ) +
  facet_wrap(.id ~ ., scales = "free_y") +
  theme_jjo() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.key = element_rect(colour = "black")
  )
```

## 2018-individuals

Let’s have a look at what’s inside `data_nes$data_2018`:
  
```{r}
# list structure
str(data_nes$year_2018, max.level = 1, give.attr = F, no.list = T)
```

> A list of `r length(data_nes$year_2018)` `data.frames`, one for each seal

For convenience, we aggregate all `r length(data_nes$year_2018)` individuals into one dataset.

```{r, eval=FALSE}
# combine all individuals
data_2018 <- rbindlist(data_nes$year_2018)

# display
DT::datatable(data_2018[sample.int(.N, 100), ], options = list(scrollX = T))
```
```{r, echo=FALSE, results='asis'}
# combine all individuals
data_2018 <- rbindlist(data_nes$year_2018)

# title
cat("<table style='width: 50%'>", paste0("<caption>", "(#tab:myDThtmltools)", "Sample of 100 random rows from `data_2018`", "</caption>"), "</table>", sep = "\n")

# display

DT::datatable(data_2018[sample.int(.N, 100), ], options = list(scrollX = T))
```

### Summary

```{r}
# raw_data
data_2018[, .(
  nb_days_recorded = uniqueN(as.Date(date)),
  nb_dives = .N,
  maxdepth_mean = mean(maxdepth),
  dduration_mean = mean(dduration),
  botttime_mean = mean(botttime),
  pdi_mean = mean(pdi, na.rm = T)
), by = .id] %>%
  sable(
    caption = "Summary diving information relative to each 2018 individual",
    digits = 2
  )
```
> Very nice dataset :)

### Some explanatory plots

#### Missing values

```{r fig.cap="Check for missing value in 2018-individuals", fig.width=9}
# build dataset to check for missing values
dataPlot <- melt(data_2018[, .(.id, is.na(.SD)), .SDcol = -c(
  ".id",
  "divenumber",
  "year",
  "month",
  "day",
  "hour",
  "min",
  "sec",
  "juldate",
  "divetype",
  "date",
  "phase",
  "lat",
  "lon"
)])
# add the id of rows
dataPlot[, id_row := c(1:.N), by = c("variable", ".id")]

# plot
ggplot(dataPlot, aes(x = variable, y = id_row, fill = value)) +
  geom_tile() +
  labs(x = "Attributes", y = "Rows") +
  scale_fill_manual(
    values = c("white", "black"),
    labels = c("Real", "Missing")
  ) +
  facet_wrap(.id ~ ., scales = "free_y") +
  theme_jjo() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.key = element_rect(colour = "black")
  )
```

So far so good, only few variables seems to have missing values:
  
```{r}
# table with percent
table_inter <- data_2018[, lapply(.SD, function(x) {
  round(length(x[is.na(x)]) * 100 / length(x), 1)
}), .SDcol = -c(
  ".id",
  "divenumber",
  "year",
  "month",
  "day",
  "hour",
  "min",
  "sec",
  "juldate",
  "divetype",
  "date",
  "phase",
  "lat",
  "lon"
)]

# find which are different from 0
cond_inter <- sapply(table_inter, function(x) {
  x == 0
})

# display the percentages that are over 0
table_inter[, which(cond_inter) := NULL] %>%
  sable(caption = "Percentage of missing values per columns having missing values!") %>%
  scroll_box(width = "100%")
```

#### Outliers {.tabset}

Ok, let's have a look at all the data. But first, we have to remove outliers. Some of them are quiet easy to spot looking at the distribution of dive duration:

##### Before {-}

```{r, fig.cap='Distribution of `dduration` for each seal. The dashed line highlight the "subjective" threshold used to remove outliers (3000 sec)', fig.height=3}
ggplot(
  data_2018[, .SD][, state := "Before"],
  aes(x = dduration, fill = .id)
) +
  geom_histogram(show.legend = FALSE) +
  geom_vline(xintercept = 3000, linetype = "longdash") +
  facet_grid(state ~ .id,
             scales = "free"
  ) +
  labs(y = "# of dives", x = "Dive duration (s)") +
  theme_jjo()
```

##### After {-}

```{r, fig.cap='Same distribution of `dduration` for each seal but after removing any `dduration` > 3000 sec. The dashed line highlight the "subjective" threshold used to remove outliers', fig.height=3}
ggplot(
  data_2018[dduration < 3000, ][][, state := "After"],
  aes(x = dduration, fill = .id)
) +
  geom_histogram(show.legend = FALSE) +
  geom_vline(xintercept = 3000, linetype = "longdash") +
  facet_grid(state ~ .id,
             scales = "free"
  ) +
  labs(x = "# of dives", y = "Dive duration (s)") +
  theme_jjo()
```

It seems much better, so let's remove any rows with `dduration` > 3000 sec.

```{r}
# filter data
data_2018_filter <- data_2018[dduration < 3000, ]

# nbrow removed
data_2018[dduration >= 3000, .(nb_row_removed = .N), by = .id] %>%
  sable(caption = "# of rows removed by 2018-individuals")
```

#### Check day and night {.tabset}

##### Light levels

```{r, fig.cap="Visualization of light level at the surface along 2018-individuals' trip", fig.height=6}
# let's first average `lightatsurf` by individuals, day since departure and hour
dataPlot <- data_2018[, .(lightatsurf = median(lightatsurf)),
                      by = .(.id, day_departure, date = as.Date(date), hour)
]

# display the result
ggplot(dataPlot, aes(x = day_departure, y = hour, fill = lightatsurf)) +
  geom_tile() +
  facet_grid(.id ~ .) +
  theme_jjo() +
  labs(x = "# of days since departure", y = "Hour", fill = "Light level at the surface") +
  theme(legend.position = c("bottom"))
```

##### Day and night detection

```{r, fig.cap="Visualization of detected night time and day time along 2018-individuals' trip", fig.height=6}
# let's first average `lightatsurf` by individuals, day since departure and hour
dataPlot <- data_2018[, .(lightatsurf = median(lightatsurf)),
                      by = .(.id, day_departure, date = as.Date(date), hour, phase)
]

# display the result
ggplot(dataPlot, aes(x = day_departure, y = hour, fill = phase)) +
  geom_tile() +
  facet_grid(.id ~ .) +
  theme_jjo() +
  labs(x = "# of days since departure", y = "Hour", fill = "Day time and night time as detected by the `cal_phase_day` function") +
  theme(legend.position = c("bottom"))
```

#### All Variables 

```{r}
names_display <- names(data_2018_filter[, -c(
  ".id",
  "date",
  "divenumber",
  "year",
  "month",
  "day",
  "hour",
  "min",
  "sec",
  "juldate",
  "divetype",
  "euphoticdepth",
  "thermoclinedepth",
  "day_departure",
  "phase",
  "lat",
  "lon"
)])
```

```{r, eval=FALSE, include=TRUE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  if (i == "maxdepth") {
    print(
      ggplot() +
        geom_point(
          data = data_2018_filter[, .(
            .id,
            date,
            thermoclinedepth
          )],
          aes(
            x = as.Date(date),
            y = -thermoclinedepth,
            colour = "Thermocline (m)"
          ),
          alpha = .2,
          size = .5
        ) +
        geom_point(
          data = data_2018_filter[, .(
            .id,
            date,
            euphoticdepth
          )],
          aes(
            x = as.Date(date),
            y = -euphoticdepth,
            colour = "Euphotic (m)"
          ),
          alpha = .2,
          size = .5
        ) +
        scale_colour_manual(
          values = c(
            "Thermocline (m)" = "red",
            "Euphotic (m)" = "black"
          ),
          name = "Zone"
        ) +
        new_scale_color() +
        geom_point(
          data = melt(data_2018_filter[, .(.id, date, get(i))], id.vars = c(".id", "date")),
          aes(
            x = as.Date(date),
            y = -value,
            col = .id
          ),
          alpha = 1 / 10,
          size = .5,
          show.legend = FALSE
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = "Maximum Depth (m)") +
        theme_jjo() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom"
        )
    )
    cat("<blockquote> Considering `ind_2018074` has slightly different values than other individuals for the thermocline depth, it would be interesting to see where the animal went. </blockquote>")
  } else if (i == "driftrate") {
    print(
      ggplot(
        data = melt(data_2018_filter[, .(.id, date, get(i), divetype)], id.vars = c(".id", "date", "divetype")),
        aes(
          x = as.Date(date),
          y = value,
          col = divetype
        )
      ) +
        geom_point(
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = "Drift Rate 'm/s", col = "Dive Type") +
        theme_jjo() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom"
        ) +
        guides(colour = guide_legend(override.aes = list(
          size = 7,
          alpha = 1
        )))
    )
  } else {
    print(
      ggplot(
        data = melt(data_2018_filter[, .(.id, date, get(i))], id.vars = c(".id", "date")),
        aes(
          x = as.Date(date),
          y = value,
          col = .id
        )
      ) +
        geom_point(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = i) +
        theme_jjo() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }
  
  cat("\n \n")
}
```

#### {.unlisted .unnumbered .tabset .tabset-fade .tabset-pills}

```{r, results='asis', cache=TRUE, echo=FALSE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  if (i == "maxdepth") {
    print(
      ggplot() +
        geom_point(
          data = data_2018_filter[, .(
            .id,
            date,
            thermoclinedepth
          )],
          aes(
            x = as.Date(date),
            y = -thermoclinedepth,
            colour = "Thermocline (m)"
          ),
          alpha = .2,
          size = .5
        ) +
        geom_point(
          data = data_2018_filter[, .(
            .id,
            date,
            euphoticdepth
          )],
          aes(
            x = as.Date(date),
            y = -euphoticdepth,
            colour = "Euphotic (m)"
          ),
          alpha = .2,
          size = .5
        ) +
        scale_colour_manual(
          values = c(
            "Thermocline (m)" = "red",
            "Euphotic (m)" = "black"
          ),
          name = "Zone"
        ) +
        new_scale_color() +
        geom_point(
          data = melt(data_2018_filter[, .(.id, date, get(i))], id.vars = c(".id", "date")),
          aes(
            x = as.Date(date),
            y = -value,
            col = .id
          ),
          alpha = 1 / 10,
          size = .5,
          show.legend = FALSE
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = "Maximum Depth (m)") +
        theme_jjo() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom"
        )
    )
    cat("<blockquote> Considering `ind_2018074` has slightly different values than other individuals for the thermocline depth, it would be interesting to see where the animal went. </blockquote>")
  } else if (i == "driftrate") {
    print(
      ggplot(
        data = melt(data_2018_filter[, .(.id, date, get(i), divetype)], id.vars = c(".id", "date", "divetype")),
        aes(
          x = as.Date(date),
          y = value,
          col = divetype
        )
      ) +
        geom_point(
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = "Drift Rate 'm/s", col = "Dive Type") +
        theme_jjo() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom"
        ) +
        guides(colour = guide_legend(override.aes = list(
          size = 7,
          alpha = 1
        )))
    )
  } else {
    print(
      ggplot(
        data = melt(data_2018_filter[, .(.id, date, get(i))], id.vars = c(".id", "date")),
        aes(
          x = as.Date(date),
          y = value,
          col = .id
        )
      ) +
        geom_point(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = i) +
        theme_jjo() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }
  
  cat("\n \n")
}
```

#### {.unlisted .unnumbered}

> Few questions, that I should look into it:
  >
  > * is the bimodal distribution of `dduration`, `desctime` due to nycthemeral migration?
  > * is the bimodal distribution of `descrate` (especially for `ind2018070` and `ind_2018072`) due to drift dive?
  > * is `lightatbott` could be used to identify bioluminescence, cause it seems there is a lot going on at the bottom?
  > * are the variations observed for `lightatsurf` is due to moon cycle?
  > * not sure why is there a bimodal distribution of `tempatbott`!
  > * `drifrate` that one is awesome! Thanks to `divetype` we can clearly see a pattern of how driftrate (and so buoyancy) change according time.
> * the bimodal distribution of `verticalspeed90` and `verticalspeed95` should be due to drift dive.

```{r, eval=FALSE, include=TRUE}
# same plot with a colored for the phase of the day
for (i in names_display) {
  cat("#####", i, "{-} \n")
  print(
    ggplot(
      data = melt(data_2018_filter[, .(.id, date, get(i), phase)],
                  id.vars = c(
                    ".id",
                    "date",
                    "phase"
                  )
      ),
      aes(
        x = as.Date(date),
        y = value,
        col = phase
      )
    ) +
      geom_point(
        alpha = 1 / 10,
        size = .5
      ) +
      facet_wrap(. ~ .id, scales = "free") +
      scale_x_date(date_labels = "%m/%Y") +
      labs(x = "Date", y = i) +
      theme_jjo() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom"
      ) +
      guides(colour = guide_legend(override.aes = list(
        size = 7,
        alpha = 1
      )))
  )
  cat("\n \n")
}
```

#### {.unlisted .unnumbered .tabset .tabset-fade .tabset-pills}

```{r, results='asis', cache=TRUE, echo=FALSE}
# same plot with a colored for the phase of the day
for (i in names_display) {
  cat("#####", i, "{-} \n")
  print(
    ggplot(
      data = melt(data_2018_filter[, .(.id, date, get(i), phase)],
                  id.vars = c(
                    ".id",
                    "date",
                    "phase"
                  )
      ),
      aes(
        x = as.Date(date),
        y = value,
        col = phase
      )
    ) +
      geom_point(
        alpha = 1 / 10,
        size = .5
      ) +
      facet_wrap(. ~ .id, scales = "free") +
      scale_x_date(date_labels = "%m/%Y") +
      labs(x = "Date", y = i) +
      theme_jjo() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom"
      ) +
      guides(colour = guide_legend(override.aes = list(
        size = 7,
        alpha = 1
      )))
  )
  cat("\n \n")
}
```


### All Variables during the first month

```{r, eval=FALSE, include=TRUE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  if (i == "maxdepth") {
    print(
      ggplot() +
        geom_point(
          data = data_2018_filter[day_departure < 32, .(
            .id,
            day_departure,
            thermoclinedepth
          )],
          aes(
            x = day_departure,
            y = -thermoclinedepth,
            colour = "Thermocline (m)",
            group = day_departure
          ),
          alpha = .2,
          size = .5
        ) +
        geom_point(
          data = data_2018_filter[day_departure < 32, .(
            .id,
            day_departure,
            euphoticdepth
          )],
          aes(
            x = day_departure,
            y = -euphoticdepth,
            colour = "Euphotic (m)",
            group = day_departure
          ),
          alpha = .2,
          size = .5
        ) +
        scale_colour_manual(
          values = c(
            "Thermocline (m)" = "red",
            "Euphotic (m)" = "black"
          ),
          name = "Zone"
        ) +
        new_scale_color() +
        geom_boxplot(
          data = melt(data_2018_filter[day_departure < 32, .(.id, day_departure, get(i))], id.vars = c(".id", "day_departure")),
          aes(
            x = day_departure,
            y = -value,
            col = .id,
            group = day_departure
          ),
          alpha = 1 / 10,
          size = .5,
          show.legend = FALSE
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        labs(x = "# days since departure", y = "Maximum Depth (m)") +
        theme_jjo() +
        theme(legend.position = "bottom")
    )
  } else {
    print(
      ggplot(
        data = melt(data_2018_filter[day_departure < 32, .(.id, day_departure, get(i))], id.vars = c(".id", "day_departure")),
        aes(
          x = day_departure,
          y = value,
          color = .id,
          group = day_departure
        )
      ) +
        geom_boxplot(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        labs(x = "# days since departure", y = i) +
        theme_jjo()
    )
  }
  cat("\n \n")
}
```

#### {.unlisted .unnumbered .tabset .tabset-fade .tabset-pills}

```{r, results='asis', cache=TRUE, echo=FALSE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  if (i == "maxdepth") {
    print(
      ggplot() +
        geom_point(
          data = data_2018_filter[day_departure < 32, .(
            .id,
            day_departure,
            thermoclinedepth
          )],
          aes(
            x = day_departure,
            y = -thermoclinedepth,
            colour = "Thermocline (m)",
            group = day_departure
          ),
          alpha = .2,
          size = .5
        ) +
        geom_point(
          data = data_2018_filter[day_departure < 32, .(
            .id,
            day_departure,
            euphoticdepth
          )],
          aes(
            x = day_departure,
            y = -euphoticdepth,
            colour = "Euphotic (m)",
            group = day_departure
          ),
          alpha = .2,
          size = .5
        ) +
        scale_colour_manual(
          values = c(
            "Thermocline (m)" = "red",
            "Euphotic (m)" = "black"
          ),
          name = "Zone"
        ) +
        new_scale_color() +
        geom_boxplot(
          data = melt(data_2018_filter[day_departure < 32, .(.id, day_departure, get(i))], id.vars = c(".id", "day_departure")),
          aes(
            x = day_departure,
            y = -value,
            col = .id,
            group = day_departure
          ),
          alpha = 1 / 10,
          size = .5,
          show.legend = FALSE
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        labs(x = "# days since departure", y = "Maximum Depth (m)") +
        theme_jjo() +
        theme(legend.position = "bottom")
    )
  } else {
    print(
      ggplot(
        data = melt(data_2018_filter[day_departure < 32, .(.id, day_departure, get(i))], id.vars = c(".id", "day_departure")),
        aes(
          x = day_departure,
          y = value,
          color = .id,
          group = day_departure
        )
      ) +
        geom_boxplot(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        labs(x = "# days since departure", y = i) +
        theme_jjo()
    )
  }
  cat("\n \n")
}
```

#### {.unlisted .unnumbered}

```{r, eval=FALSE, include=TRUE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  print(
    ggplot(
      data = melt(data_2018_filter[
        day_departure < 32,
        .(.id, day_departure, get(i), phase)
      ],
      id.vars = c(".id", "day_departure", "phase")
      ),
      aes(
        x = day_departure,
        y = value,
        color = phase,
        group = interaction(day_departure, phase),
      )
    ) +
      geom_boxplot(
        alpha = 1 / 10,
        size = .5
      ) +
      facet_wrap(. ~ .id, scales = "free") +
      labs(x = "# days since departure", y = i) +
      theme_jjo() +
      theme(legend.position = "bottom")
  )
  cat("\n \n")
}
```

#### {.unlisted .unnumbered .tabset .tabset-fade .tabset-pills}

```{r, results='asis', cache=TRUE, echo=FALSE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  print(
    ggplot(
      data = melt(data_2018_filter[
        day_departure < 32,
        .(.id, day_departure, get(i), phase)
      ],
      id.vars = c(".id", "day_departure", "phase")
      ),
      aes(
        x = day_departure,
        y = value,
        color = phase,
        group = interaction(day_departure, phase),
      )
    ) +
      geom_boxplot(
        alpha = 1 / 10,
        size = .5
      ) +
      facet_wrap(. ~ .id, scales = "free") +
      labs(x = "# days since departure", y = i) +
      theme_jjo() +
      theme(legend.position = "bottom")
  )
  cat("\n \n")
}
```

#### Correlation

Can we find nice correlation?
  
```{r, fig.cap="Correlation matrix (crosses indicate non significant correlation)", fig.width=10, fig.height=10}
# compute correlation
corr_2018 <- round(cor(data_2018_filter[, names_display, with = F],
                       use = "pairwise.complete.obs"
), 1)

# replace NA value by 0
corr_2018[is.na(corr_2018)] <- 0

# compute p_values
corr_p_2018 <- cor_pmat(data_2018_filter[, names_display, with = F])

# replace NA value by 0
corr_p_2018[is.na(corr_p_2018)] <- 1

# display
ggcorrplot(
  corr_2018,
  p.mat = corr_p_2018,
  hc.order = TRUE,
  method = "circle",
  type = "lower",
  ggtheme = theme_jjo(),
  sig.level = 0.05,
  colors =  c("#00AFBB", "#E7B800", "#FC4E07")
)
```

Another way to see it:
  
```{r}
# flatten correlation matrix
cor_result_2018 <- flat_cor_mat(corr_2018, corr_p_2018)

# keep only the one above .7
cor_result_2018[cor >= .7, ][order(-abs(cor))] %>%
  sable(caption = "Pairwise correlation above 0.75 and associated p-values")
```

> I guess nothing unexpected here, I'll have to check with Patrick about the efficiency ;)

### Dive Type

```{r, fig.cap="Proportion dive types"}
# dataset to plot proportional area plot
data_2018_filter[, sum_id := .N, by = .(.id, day_departure)][, sum_id_days := .N, by = .(.id, day_departure, divetype)][, prop := sum_id_days /
                                                                                                                          sum_id]
dataPlot <- unique(data_2018_filter[, .(prop, .id, divetype, day_departure)])

# area plot
ggplot(dataPlot, aes(
  x = as.numeric(day_departure),
  y = prop,
  fill = as.character(divetype)
)) +
  geom_area(alpha = 0.6, size = 1) +
  facet_wrap(.id ~ ., scales = "free") +
  theme_jjo() +
  theme(legend.position = "bottom") +
  labs(x = "# of days since departure", y = "Proportion of dives", fill = "Dive types")
```

### Dive duration *vs.* Maximum depth {.tabset}

#### Colored by ID

```{r, fig.cap="Dive duration vs. Maximum Depth colored 2018-individuals"}
# plot
ggplot(data = data_2018_filter, aes(y = dduration, x = maxdepth, col = .id)) +
  geom_point(size = .5, alpha = .1, show.legend = FALSE) +
  facet_wrap(.id ~ .) +
  labs(x = "Maximum depth (m)", y = "Dive duration (s)") +
  theme_jjo()
```

#### Colored by Dive Type

```{r, fig.cap="Dive duration vs. Maximum Depth colored by Dive Type"}
# plot
ggplot(data = data_2018_filter, aes(y = dduration, x = maxdepth, col = divetype)) +
  geom_point(size = .5, alpha = .1) +
  facet_wrap(.id ~ .) +
  guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))) +
  labs(x = "Maximum depth (m)", y = "Dive duration (s)") +
  theme_jjo() +
  theme(legend.position = "bottom")
```

#### Colored by # days since departure

```{r, fig.cap="Dive duration vs. Maximum Depth colored by # days since departure"}
# plot
ggplot(data = data_2018_filter[, prop_track := (day_departure * 100) / max(day_departure), by = .id], aes(y = dduration, x = maxdepth, col = prop_track)) +
  geom_point(size = .5, alpha = .1) +
  facet_wrap(.id ~ .) +
  labs(x = "Maximum depth (m)", y = "Dive duration (s)", col = "Proportion of completed track (%)") +
  scale_color_continuous(type = "viridis") +
  theme_jjo() +
  theme(legend.position = "bottom")
```

#### Colored by phases of day

```{r, fig.cap="Dive duration vs. Maximum Depth colored by phases of the day"}
# plot
ggplot(data = data_2018_filter, aes(y = dduration, x = maxdepth, col = phase)) +
  geom_point(size = .5, alpha = .1) +
  facet_wrap(.id ~ .) +
  guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))) +
  labs(x = "Maximum depth (m)", y = "Dive duration (s)", col = "Phases of the day") +
  theme_jjo() +
  theme(legend.position = "bottom")
```

> There seems to be a *patch* for high depths (especially visible for `ind2018070`), but I don't know what it could be linked to...

### Drift Rate

> In the following graphs:
>
> * `driftrate` is calculated using only `divetype == "2: drift"`
> * whereas all the others variables are calculated all dives considered

```{r}
# build dataset
dataPlot <- data_2018_filter[divetype == "2: drift",
                             # median drift rate for drift dive
                             .(driftrate = median(driftrate, na.rm = T)),
                             by = .(.id, day_departure)
][data_2018_filter[,
                   .(
                     # median dive duration all dives considered
                     dduration = median(dduration, na.rm = T),
                     # median max depth all dives considered
                     maxdepth = median(maxdepth, na.rm = T),
                     # median bottom dives all dives considered
                     botttime = median(botttime, na.rm = T)
                   ),
                   by = .(.id, day_departure)
],
on = c(".id", "day_departure")
]
```

```{r, fig.cap="Drift rate vs. Bottom time"}
# plot
ggplot(dataPlot, aes(x = botttime, y = driftrate, col = .id)) +
  geom_point(size = .5, alpha = .5) +
  geom_smooth(method = "lm") +
  guides(color = FALSE) +
  facet_wrap(.id ~ .) +
  scale_x_continuous(limits = c(0, 700)) +
  labs(x = "Daily median Bottom time (s)", y = "Daily median drift rate (m.s-1)") +
  theme_jjo()
```

```{r, fig.cap="Drift rate vs. Maximum depth"}
# plot
ggplot(dataPlot, aes(x = maxdepth, y = driftrate, col = .id)) +
  geom_point(size = .5, alpha = .5) +
  geom_smooth(method = "lm") +
  guides(color = FALSE) +
  facet_wrap(.id ~ .) +
  labs(x = "Daily median Maximum depth (m)", y = "Daily median drift rate (m.s-1)") +
  theme_jjo()
```

```{r, fig.cap="Drift rate vs. Dive duration"}
# plot
ggplot(dataPlot, aes(x = dduration, y = driftrate, col = .id)) +
  geom_point(size = .5, alpha = .5) +
  geom_smooth(method = "lm") +
  guides(color = FALSE) +
  facet_wrap(.id ~ .) +
  labs(x = "Daily median Dive duration (s)", y = "Daily median drift rate (m.s-1)") +
  theme_jjo()
```

### GPS data

```{r, fig.cap="Map with polylines", fig.width=8, eval=FALSE}
# This piece of code is only there to show how to draw a polylines with a gradient color using leaflet.
# We're not using it due to the size of the created map, and will continue using circle marker

# datasets used to display map
df_driftrate = unique(data_2018_filter[.id == "ind_2018070" & 
                                         divetype == "2: drift", 
                                       .(.id, lat, lon, dduration)])

# color palette
pal <- colorNumeric(
  palette = "YlGnBu",
  domain = df_driftrate$dduration
)

# add 
df_driftrate[, `:=`(nextLat = shift(lat),
                    nextLon = shift(lon),
                    color = pal(df_driftrate$dduration))]

# interactive map
gradient_map <- leaflet() %>%
  setView(lng = -122, lat = 38, zoom = 2) %>%
  addTiles() 

# add lines
for (i in 1:nrow(df_driftrate)) {
  gradient_map <- addPolylines(
    map = gradient_map,
    data = df_driftrate,
    lat = as.numeric(df_driftrate[i, c('lat', 'nextLat')]),
    lng = as.numeric(df_driftrate[i, c('lon', 'nextLon')]),
    color = df_driftrate[i, color],
    weight = 3,
    group = "individual_1"
  )
}

# add layer control
gradient_map <- addLayersControl(
  map = gradient_map,
  overlayGroups = c("individual_1"),
  options = layersControlOptions(collapsed = FALSE)
)

# format(object.size(gradient_map), units = "Mb")
```

Because for some data the contrast in changes was not enough marked, the only treatment applied on these data is to remove outliers for each variable using the interquartile range rule. 

```{r}
# interactive map
gradient_map <- leaflet() %>%
  setView(lng = -132, lat = 48, zoom = 4) %>%
  addTiles()

# loop by individuals and variable
grps = NULL
for (i in seq(data_2018_filter[!is.na(lat),unique(.id)])){
  for (k in c("dduration","maxdepth","efficiency", "driftrate")){
    if (k == "driftrate"){
      # set dataset used to plot
      dataPlot = unique(data_2018_filter %>% 
                          .[order(date),] %>% 
                          .[.id==data_2018_filter[!is.na(lat), unique(.id)][i] &
                              divetype == "2: drift" &
                              !is.na(get(k)), 
                            c("lat", "lon", k), with=FALSE] %>% 
                          .[!is_outlier(get(k)),])
      # color palette creation
      colPal <- colorNumeric(
        palette = "BrBG",
        domain = seq(-dataPlot[,max(abs(driftrate))],
                     dataPlot[,max(abs(driftrate))],
                     0.1)
      )
    } else {
      # set dataset used to plot
      dataPlot = unique(data_2018_filter %>% 
                          .[order(date),] %>% 
                          .[.id==data_2018_filter[!is.na(lat), unique(.id)][i] & 
                              divetype != "2: drift" &
                              !is.na(get(k)), 
                            c("lat", "lon", k), with=FALSE] %>% 
                          .[!is_outlier(get(k)),]) 
      # color palette creation
      colPal <- colorNumeric(
        palette = "YlGnBu",
        domain = dataPlot[,get(k)]
      )
    }
    
    # add color to dataset
    dataPlot[, color := colPal(dataPlot[,get(k)])]
    # add size column
    dataPlot[, radius := 3]
    # mark the beginning of the trip
    dataPlot[1, `:=` (color = "green",
                      radius = 4)]
    # mark the end of the trip
    dataPlot[.N, `:=` (color = "red",
                       radius = 4)]
    # reorder to make the end and the beginning in front
    dataPlot = rbind(dataPlot[-1,],dataPlot[1,])
    # add markers to map
    gradient_map <- addCircleMarkers(
      map = gradient_map,
      data = dataPlot,
      lat = ~lat,
      lng = ~lon,
      radius = ~radius,
      stroke = FALSE,
      color = ~color,
      fillOpacity = 1,
      group = paste(data_2018_filter[,unique(.id)][i], "-", k)
    ) %>% 
      addLegend("bottomleft", 
                data = dataPlot,
                group = paste(data_2018_filter[,unique(.id)][i], "-", k),
                pal = colPal, 
                values = ~get(k),
                title = k,
                opacity = 1
      )
    # retrieve groups
    grps = c(grps, paste(data_2018_filter[,unique(.id)][i], "-", k))
  }
}

# add layer control
gradient_map <- addLayersControl(
  map = gradient_map,
  overlayGroups = grps,
  options = layersControlOptions(collapsed = TRUE)
) %>% hideGroup(grps)
```

```{r, fig.cap="Tracking data 2018 individuals (green and red dot respectively indicate the beginning and the end of each trip)", fig.width=8}
# display
gradient_map
```

