---
title: "Data Exploration SES - 2014"
author: "Joffrey JOUMAA"
date: "`r invisible(Sys.setlocale(locale = 'C')); format(Sys.Date(), format = '%B %d, %Y')`"
output:
  bookdown::html_document2:
    css: cosmo_custom.css
    number_sections: yes
    code_folding: show
    df_print: default
    fig_caption: yes
    code_download: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
vignette: >
  %\VignetteIndexEntry{Data Exploration SES - 2014}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# command to build package without getting vignette error
# https://github.com/rstudio/renv/issues/833
# devtools::check(build_args=c("--no-build-vignettes"))

# reduce png size
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

# global option relative to rmarkdown
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # tidy = TRUE,
  cache.lazy = FALSE,
  optipng = "-o7 -quiet",
  pngquant = "--speed=1"
)

# library
library(data.table)
library(magrittr)
library(kableExtra)

# remove some warnings
suppressWarnings(library(ggplot2))

# define my own table format: https://github.com/haozhu233/kableExtra/issues/374
sable <- function(x, escape = T, ...) {
  knitr::kable(x, escape = escape, ...) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "responsive"),
      full_width = F
    )
}

# theme ggplot
# based: https://benjaminlouis-stat.fr/en/blog/2020-05-21-astuces-ggplot-rmarkdown/
theme_jjo <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # the whole figure
      # plot.title = element_text(size = rel(1), face = "bold", margin = margin(0,0,5,0), hjust = 0),
      # figure area
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      # axes
      # axis.title = element_text(size = rel(0.85), face = "bold"),
      # axis.text = element_text(size = rel(0.70), face = "bold"),
      axis.line = element_line(color = "black", arrow = arrow(length = unit(0.2, "lines"), type = "closed")),
      # legend
      # legend.title = element_text(size = rel(0.85), face = "bold"),
      # legend.text = element_text(size = rel(0.70), face = "bold"),
      # legend.key = element_rect(fill = "transparent", colour = NA),
      # legend.key.size = unit(1.5, "lines"),
      # legend.background = element_rect(fill = "transparent", colour = NA),
      # Les <U+00E9>tiquettes dans le cas d'un facetting
      strip.background = element_rect(fill = "#888888", color = "#888888"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "white", margin = margin(5, 0, 5, 0))
    )
}
```

This document aims at exploring the dataset of 9 southern-individuals in 2014. For that purpose, we need first to load the `weanlingNES` package to load data.

```{r data-exploration-2018-1}
# load library
library(weanlingNES)

# load data
# data("data_ses", package = "weanlingNES")
load("../data/data_ses.rda")
```

Let’s have a look at what’s inside `data_ses$data_2014`:
  
```{r data-exploration-2018-2}
# list structure
str(data_ses$year_2014, max.level = 1, give.attr = F, no.list = T)
```

> A list of `r length(data_ses$year_2014)` `data.frames`, one for each seal

For convenience, we aggregate all `r length(data_ses$year_2014)` individuals into one dataset.

```{r data-exploration-2018-3, eval=FALSE}
# combine all individuals
data_2014 <- rbindlist(data_ses$year_2014)

# display
DT::datatable(data_2014[sample.int(.N, 10), ], options = list(scrollX = T))
```
```{r data-exploration-2018-4, echo=FALSE, results='asis'}
# combine all individuals
data_2014 <- rbindlist(data_ses$year_2014)

# title
cat("<table style='width: 50%'>",
  paste0(
    "<caption>",
    "(#tab:myDThtmltools)",
    "Sample of 10 random rows from `data_2014`",
    "</caption>"
  ),
  "</table>",
  sep = "\n"
)

# display
DT::datatable(data_2014[sample.int(.N, 10), ], options = list(scrollX = T))
```

## Summary

```{r data-exploration-2018-5}
# raw_data
data_2014[, .(
  nb_days_recorded = uniqueN(as.Date(date)),
  nb_dives = .N,
  maxdepth_mean = mean(maxdepth),
  dduration_mean = mean(dduration),
  botttime_mean = mean(botttime),
  pdi_mean = mean(pdi, na.rm = T)
), by = .id] %>%
  sable(
    caption = "Summary diving information relative to each 2018 individual",
    digits = 2
  )
```

## Some explanatory plots

### Missing values

```{r data-exploration-2018-6, fig.cap="Check for missing value in 2018-individuals", fig.width=9}
# build dataset to check for missing values
dataPlot <- melt(data_2014[, .(.id, is.na(.SD)), .SDcol = -c(
  ".id",
  "divenumber",
  "year",
  "month",
  "day",
  "hour",
  "min",
  "sec",
  "juldate",
  "divetype",
  "date"
  # "phase",
  # "lat",
  # "lon"
)])
# add the id of rows
dataPlot[, id_row := c(1:.N), by = c("variable", ".id")]

# plot
ggplot(dataPlot, aes(x = variable, y = id_row, fill = value)) +
  geom_tile() +
  labs(x = "Attributes", y = "Rows") +
  scale_fill_manual(
    values = c("white", "black"),
    labels = c("Real", "Missing")
  ) +
  facet_wrap(.id ~ ., scales = "free_y") +
  theme_jjo() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.key = element_rect(colour = "black")
  )
```

Let's look closer to the variables with missing values:

```{r}
# table with percent
table_inter <- data_2014[, lapply(.SD, function(x) {
  round(length(x[is.na(x)]) * 100 / length(x), 1)
}), .SDcol = -c(
  ".id",
  "divenumber",
  "year",
  "month",
  "day",
  "hour",
  "min",
  "sec",
  "juldate",
  "divetype",
  "date"
  # "phase",
  # "lat",
  # "lon"
)]

# find which are different from 0
cond_inter <- sapply(table_inter, function(x) {
  x == 0
})

# display the percentages that are over 0
table_inter[, which(cond_inter) := NULL] %>%
  sable(caption = "Percentage of missing values per columns having missing values!") %>%
  scroll_box(width = "100%")
```

Nothing bad, missing values seem to occur only in column we are not interested in.

### Outliers

Let’s see if we have some outliers. Some of them are quiet easy to spot looking at the distribution of dive duration:

```{r fig.cap="Distribution of `dduration`, `maxdepth` and `driftrate` for each seal", fig.show = "hold", fig.height=1.5}
# plot that, weirdly, doesn't free x axis...
# ggplot(
#   melt(data_2014,
#        id.vars = c(".id"),
#        measure.vars = c("dduration","maxdepth","driftrate")),
#   aes(x = value, fill = .id)
# ) +
#   geom_histogram(show.legend = FALSE) +
#   facet_grid(variable ~ .id,
#              scales = "free"
#   ) +
#   labs(y = "# of dives") +
#   theme_jjo()
ggplot(
  melt(data_2014,
    id.vars = c(".id"),
    measure.vars = c("dduration")
  ),
  aes(x = value, fill = .id)
) +
  geom_histogram(show.legend = FALSE) +
  facet_grid(variable ~ .id,
    scales = "free"
  ) +
  labs(y = "# of dives") +
  theme_jjo() +
  theme(
    axis.title.x = element_blank(),
    text = element_text(size = 8)
  )
ggplot(
  melt(data_2014,
    id.vars = c(".id"),
    measure.vars = c("maxdepth")
  ),
  aes(x = value, fill = .id)
) +
  geom_histogram(show.legend = FALSE) +
  facet_grid(variable ~ .id,
    scales = "free"
  ) +
  labs(y = "# of dives") +
  theme_jjo() +
  theme(
    strip.text.x = element_blank(),
    axis.title.x = element_blank(),
    text = element_text(size = 8)
  )
ggplot(
  melt(data_2014,
    id.vars = c(".id"),
    measure.vars = c("driftrate")
  ),
  aes(x = value, fill = .id)
) +
  geom_histogram(show.legend = FALSE) +
  facet_grid(variable ~ .id,
    scales = "free"
  ) +
  labs(y = "# of dives") +
  theme_jjo() +
  theme(
    strip.text.x = element_blank(),
    text = element_text(size = 8)
  )
```

Nothing that obvious, which is great :)

### All Variables

```{r}
names_display <- names(data_2014[, -c(
  ".id",
  "date",
  "divenumber",
  "year",
  "month",
  "day",
  "hour",
  "min",
  "sec",
  "juldate",
  "divetype",
  "day_departure",
  "phase",
  "lat",
  "lon",
  "sp"
)])

# calulate the median of driftrate for each day
median_driftrate <- data_2014[divetype == "2: drift",
  .(driftrate = quantile(driftrate, 0.5)),
  by = .(date = as.Date(date), .id)
]

# let's identity when the smooth changes sign
changes_driftrate <- median_driftrate %>%
  .[, .(
    y_smooth = predict(loess(driftrate ~ as.numeric(date), span = 0.25)),
    date
  ), by = .id] %>%
  .[c(FALSE, diff(sign(y_smooth)) != 0), ]
```

#### Full trip duration

```{r eval=FALSE, include=TRUE}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  if (i == "driftrate") {
    print(
      ggplot(
        data = melt(data_2014[, .(.id, date, get(i), divetype)],
          id.vars = c(".id", "date", "divetype")
        ),
        aes(
          x = as.Date(date),
          y = value,
          col = divetype
        )
      ) +
        geom_point(
          alpha = 1 / 10,
          size = .5
        ) +
        geom_vline(
          data = changes_driftrate,
          aes(xintercept = date),
          linetype = 2
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = "Drift Rate m/s", col = "Dive Type") +
        theme_jjo() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom"
        ) +
        guides(colour = guide_legend(override.aes = list(
          size = 7,
          alpha = 1
        )))
    )
  } else {
    print(
      ggplot(
        data = melt(data_2014[, .(.id, date, get(i))],
          id.vars = c(".id", "date")
        ),
        aes(
          x = as.Date(date),
          y = value,
          col = .id
        )
      ) +
        geom_point(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        geom_vline(
          data = changes_driftrate,
          aes(xintercept = date),
          linetype = 2
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = i) +
        theme_jjo() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }

  cat("\n \n")
}
```

The vertical dashed lines represent changes in buoyancy (see `vignette("buoyancy_detect")` for more information)

#### {.unlisted .unnumbered .tabset .tabset-fade .tabset-pills}

```{r results='asis', cache=TRUE, echo=FALSE, fig.height=7}
for (i in names_display) {
  cat("#####", i, "{.unlisted .unnumbered} \n")
  if (i == "driftrate") {
    print(
      ggplot(
        data = melt(data_2014[, .(.id, date, get(i), divetype)],
          id.vars = c(".id", "date", "divetype")
        ),
        aes(
          x = as.Date(date),
          y = value,
          col = divetype
        )
      ) +
        geom_point(
          alpha = 1 / 10,
          size = .5
        ) +
        geom_vline(
          data = changes_driftrate,
          aes(xintercept = date),
          linetype = 2
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = "Drift Rate m/s", col = "Dive Type") +
        theme_jjo() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom"
        ) +
        guides(colour = guide_legend(override.aes = list(
          size = 7,
          alpha = 1
        )))
    )
  } else {
    print(
      ggplot(
        data = melt(data_2014[, .(.id, date, get(i))],
          id.vars = c(".id", "date")
        ),
        aes(
          x = as.Date(date),
          y = value,
          col = .id
        )
      ) +
        geom_point(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        geom_vline(
          data = changes_driftrate,
          aes(xintercept = date),
          linetype = 2
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        scale_x_date(date_labels = "%m/%Y") +
        labs(x = "Date", y = i) +
        theme_jjo() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }

  cat("\n \n")
}
```

#### First month at sea

```{r data-exploration-2018-18, eval=FALSE, include=TRUE}
for (i in names_display) {
  # subtitle
  cat("#####", i, "{.unlisted .unnumbered} \n")
  
  # print plot
  print(
      ggplot(
        data = melt(data_2014[day_departure < 32, 
                                     .(.id, day_departure, get(i))], 
                    id.vars = c(".id", "day_departure")),
        aes(
          x = day_departure,
          y = value,
          color = .id,
          group = day_departure
        )
      ) +
        geom_boxplot(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        labs(x = "# days since departure", y = i) +
        theme_jjo()
    )
  cat("\n \n")
}
```

#### {.unlisted .unnumbered .tabset .tabset-fade .tabset-pills}

```{r data-exploration-2018-19, results='asis', cache=TRUE, echo=FALSE}
for (i in names_display) {
  # subtitle
  cat("#####", i, "{.unlisted .unnumbered} \n")
  
  # print plot
  print(
      ggplot(
        data = melt(data_2014[day_departure < 32, 
                                     .(.id, day_departure, get(i))], 
                    id.vars = c(".id", "day_departure")),
        aes(
          x = day_departure,
          y = value,
          color = .id,
          group = day_departure
        )
      ) +
        geom_boxplot(
          show.legend = FALSE,
          alpha = 1 / 10,
          size = .5
        ) +
        facet_wrap(. ~ .id, scales = "free") +
        labs(x = "# days since departure", y = i) +
        theme_jjo()
    )
  cat("\n \n")
}
```

## Dive Type

```{r fig.cap = "Evolution of dive type proportion"}
# dataset to plot proportional area plot
data_2014[, sum_id := .N, by = .(.id, day_departure)] %>%
  .[, sum_id_days := .N, by = .(.id, day_departure, divetype)] %>%
  .[, prop := sum_id_days / sum_id]
dataPlot <- unique(data_2014[, .(prop, .id, divetype, day_departure)])

# area plot
ggplot(dataPlot, aes(
  x = as.numeric(day_departure),
  y = prop,
  fill = as.character(divetype)
)) +
  geom_area(alpha = 0.6, size = 1) +
  facet_wrap(.id ~ ., scales = "free") +
  theme_jjo() +
  theme(legend.position = "bottom") +
  labs(x = "# of days since departure", 
       y = "Proportion of dives", 
       fill = "Dive types")
```

## Drift Rate

> In the following graphs:
>
> * `driftrate` is calculated using only `divetype == "2: drift"`
> * whereas all the others variables are calculated all dives considered
> * Dives were `driftrate > 0` were excluded

```{r data-exploration-2018-29}
# build dataset
dataPlot <- data_2014[divetype == "2: drift" &
                        driftrate < 0,
                      # median drift rate for drift dive
                      .(driftrate = median(driftrate, na.rm = T)),
                      by = .(.id, day_departure)] %>%
  # merge to get other parameters including all dives
  .[data_2014[driftrate < 0,
              .(
                # median dive duration all dives considered
                dduration = median(dduration, na.rm = T),
                # median max depth all dives considered
                maxdepth = median(maxdepth, na.rm = T),
                # median bottom dives all dives considered
                botttime = median(botttime, na.rm = T)
              ),
              by = .(.id, day_departure)],
    on = c(".id", "day_departure")]
```

```{r data-exploration-2018-30, fig.cap="Drift rate vs. Bottom time"}
# plot
ggplot(dataPlot, aes(x = botttime, y = driftrate, col = .id)) +
  geom_point(size = .5, alpha = .5) +
  geom_smooth(method = "lm") +
  guides(color = "none") +
  facet_wrap(.id ~ .) +
  scale_x_continuous(limits = c(0, 700)) +
  labs(x = "Daily median Bottom time (s)", 
       y = "Daily median drift rate (m.s-1)") +
  theme_jjo()
```

```{r data-exploration-2018-31, fig.cap="Drift rate vs. Maximum depth"}
# plot
ggplot(dataPlot, aes(x = maxdepth, y = driftrate, col = .id)) +
  geom_point(size = .5, alpha = .5) +
  geom_smooth(method = "lm") +
  guides(color = "none") +
  facet_wrap(.id ~ .) +
  labs(x = "Daily median Maximum depth (m)", 
       y = "Daily median drift rate (m.s-1)") +
  theme_jjo()
```

```{r data-exploration-2018-32, fig.cap="Drift rate vs. Dive duration"}
# plot
ggplot(dataPlot, aes(x = dduration, y = driftrate, col = .id)) +
  geom_point(size = .5, alpha = .5) +
  geom_smooth(method = "lm") +
  guides(color = "none") +
  facet_wrap(.id ~ .) +
  labs(x = "Daily median Dive duration (s)", 
       y = "Daily median drift rate (m.s-1)") +
  theme_jjo()
```

## Behavioral Aerobic Dive Limit (bADL)

Based on [Shero et *al.* (2018)](https://www.researchgate.net/publication/222683042_To_breathe_or_not_to_breathe_Optimal_breathing_aerobic_dive_limit_and_oxygen_stores_in_deep-diving_blue-eyed_shags), we decided to look at the *bADL* as the 95th percentile of dive duration each day, for those with $n \geq 8$. This threshold was chosen following this figure, please note that this number is particularly low cause only one dive every 2.2 hours was sampled:

```{r data-exploration-2018-36, fig.cap="Distribution of the number of dives each day. The threshold used to calculate bADL is fixed at 50 dives per day.", fig.height=3}
ggplot(data_2014[,.(nb_dives = .N), 
                        by = .(.id, day_departure)], 
       aes(x=nb_dives, fill=.id)) +
  geom_histogram(show.legend = FALSE) + 
  facet_wrap(.~.id) +
  labs(y="# of days", x = "# of dives per day") +
  theme_jjo() +
  theme(text = element_text(size = 8))
```

```{r data-exploration-2018-37, fig.cap="Behavioral ADL vs. drift rate along animals' trip (Am I the only one seeing some kind of relationship?)"}
# select day that have at least 50 dives
days_to_keep = data_2014[,
                                .(nb_dives = .N),
                                by = .(.id, day_departure)] %>%
  .[nb_dives >= 8,]

# keep only those days
data_2014_complete_day = merge(data_2014,
                                      days_to_keep,
                                      by = c(".id", "day_departure"))

# data plot
dataPlot = data_2014_complete_day[divetype=="1: foraging",
                                         .(badl = quantile(dduration, 0.95)),
                                         by = .(.id, day_departure)]

# combine two datasets to be able to use a second axis
# https://stackoverflow.com/questions/49185583/two-y-axes-with-different-scales-for-two-datasets-in-ggplot2
dataMegaPlot = rbind(data_2014_complete_day[divetype == "2: drift"] %>%
                       .[, .(w = .id,
                             y = driftrate,
                             x = day_departure,
                             z = "second_plot")],
                     dataPlot[, .(
                       w = .id,
                       # tricky one
                       y = (badl / 1000) - 1,
                       x = day_departure,
                       z = "first_plot"
                     )])

# plot
ggplot() +
  geom_point(
    data = dataMegaPlot[z == "second_plot", ],
    aes(x = x, y = y),
    alpha = 1 / 10,
    size = 0.5,
    color = "grey40",
    show.legend = FALSE
  ) +
  geom_path(data = dataMegaPlot[z == "first_plot", ],
            aes(x = x, y = y, color = w),
            show.legend = FALSE) +
  scale_y_continuous(
    # Features of the first axis
    name = "Drift rate (m/s)",
    # Add a second axis and specify its features
    sec.axis = sec_axis( ~ (. * 1000) + 1000, 
                         name = "Behavioral Aerobic Dive Limit (s)")
  ) +
  labs(x = "# days since departure") +
  facet_wrap(w ~ .) +
  theme_jjo()
```

> Looking at this graph, I want to believe that there is some kind of relationship between the *bADL* as defined by [Shero et *al.* (2018)](https://www.researchgate.net/publication/222683042_To_breathe_or_not_to_breathe_Optimal_breathing_aerobic_dive_limit_and_oxygen_stores_in_deep-diving_blue-eyed_shags) and the drift rate (and so buyoancy).

```{r data-exploration-2018-38}
# get badl
dataplot_1 = data_2014_complete_day[,
                              .(badl = quantile(dduration, 0.95)),
                              by = .(.id, day_departure)]
# get driftrate
dataplot_2 = data_2014_complete_day[divetype == "2: drift",
                              .(driftrate = median(driftrate)),
                              by = .(.id, day_departure)]

# merge
dataPlot = merge(dataplot_1,
                 dataplot_2,
                 by = c(".id", "day_departure"),
                 all = TRUE)

# plot
ggplot(data = dataPlot[driftrate < 0, ], 
       aes(x = badl, y = driftrate, col = .id)) +
  geom_point(show.legend = FALSE) +
  geom_smooth(method = "lm", show.legend = FALSE) +
  facet_wrap(.id~., scales = "free") +
  labs(x = "Behavioral Aerobic Dive Limit (s)",
       y = "Drift rate (m/s)") +
  theme_jjo()
```
