---
title: "How to deal with netCDF files?"
author: "Joffrey JOUMAA"
date: "`r invisible(Sys.setlocale(locale = 'C')); format(Sys.Date(), format = '%B %d, %Y')`"
output:
  bookdown::html_document2:
    css: cosmo_custom.css
    number_sections: yes
    code_folding: show
    df_print: default
    fig_caption: yes
    code_download: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
vignette: >
  %\VignetteIndexEntry{How to deal with netCDF files?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# command to build package without getting vignette error
# https://github.com/rstudio/renv/issues/833
# devtools::check(build_args=c("--no-build-vignettes"))

# # reduce png size
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
# knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

# global option relative to rmarkdown
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # tidy = TRUE,
  optipng = "-o7 -quiet",
  pngquant = "--speed=1"
)

# library
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(magrittr)
library(gganimate)
library(weanlingNES)

# remove some warnings
suppressWarnings(library(ggplot2))

# define my own table format: https://github.com/haozhu233/kableExtra/issues/374
sable <- function(x, escape = T, ...) {
  knitr::kable(x, escape = escape, ...) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "responsive"),
      full_width = F
    )
}
```

## What's a netCDF file?

> NetCDF (Network Common Data Form) is a set of software libraries and self-describing, machine-independent data formats that support the creation, access, and sharing of array-oriented scientific data. The project homepage[^1] is hosted by the Unidata program at the University Corporation for Atmospheric Research (UCAR). They are also the chief source of netCDF software, standards development, updates, etc. The format is an open standard. NetCDF Classic and 64-bit Offset Format are an international standard of the Open Geospatial Consortium.[^2] 
>
> -- <cite>[Wikipedia](https://en.wikipedia.org/wiki/NetCDF)</cite>

The way I see a netCDF file is that it's a matrix that usually contains environmental data across time and location, plus extra information (called *metadata*) such as where does the dataset comes from, or its resolution.

![Here is an example[^3] of the structure of a netCDF file containing information on temperature and pressure accross time for several location, but also on the elevation and the land cover](https://web.itu.edu.tr/~tokerem/Presentation/netcdf_1.png)

## How to open it

Super simple, using the [`tidync`](https://cran.r-project.org/web/packages/ncdf4/index.html) package! Let's say you have a netcdf file called **global-reanalysis-phy-001-031-grepv2-daily_1639765258612.nc**, here is what you have to do to import this file:

```{r netcdf-read-1}
# first load `tidync` package
library(tidync)

# then import your file
df_nc <- tidync("../inst/extdata/copernicus/global-reanalysis-phy-001-031-grepv2-daily_1639765258612.nc") 

# flatten the multidimensional array
df = df_nc %>% hyper_tibble() %>% 
  # not mandatory, but here is the code to convert to a data.table
  setDT()
```
```{r netcdf-read-2}
# print the first 10 rows
df[sample(nrow(df), 10),]  %>%
  sable(
    caption = "First 10 rows of the flatten netCDF file",
    digits = 2
  )
```

Now we can see that the time is not in a proper format. The first thing to do is to look into the metadata to see if we could find some kind of explanation. To do that, we're going to use the `ncmeta` package, developped for this purpose.

```{r netcdf-read-3}
# load ncmeta library
library(ncmeta)

# then we use the function nc_atts to access attributes of times
print(
  nc_atts("../inst/extdata/copernicus/global-reanalysis-phy-001-031-grepv2-daily_1639765258612.nc", 
          "time") %>% 
    # we want information on the unit for time column 
    dplyr::filter(name == "units") %>% 
    pull(value)
)
```
The column `time` seems to be the number of days since `1950-01-01`. Let's use that information to format the `time` column in a proper way.

```{r netcdf-read-4}
# convert time into a readable format
df[, time := as.Date(time, origin=as.Date("1950-01-01"))]

# print
df[sample(nrow(df), 10),]  %>%
  sable(
    caption = "First 10 rows of the flatten netCDF file, with reformat time column",
    digits = 2
  )
```

## How to visualize it

Well there are plenty of ways to display data within a netCDF. Here, we are only presenting one simple way to do it using `ggplot2` and `gganimate` package.

```{r netcdf-read-5}
# first let's calculate the velocity norm
df[, `:=` (vel_cglo = sqrt(uo_cglo ^ 2 + vo_cglo ^ 2),
           vel_oras = sqrt(uo_oras ^ 2 + vo_oras ^ 2),
           vel_foam = sqrt(uo_foam ^ 2 + vo_foam ^ 2),
           vel_glor = sqrt(uo_glor ^ 2 + vo_glor ^ 2),
           month = month(time),
           year = year(time))]

# average by month
df_summarize = df[, .(vel_cglo = mean(vel_cglo, na.rm = T),
                      vel_oras = mean(vel_oras, na.rm = T),
                      vel_foam = mean(vel_foam, na.rm = T),
                      vel_glor = mean(vel_glor, na.rm = T)),
                  by = .(date = paste(year, '-', month),
                         longitude,
                         latitude)] %>%
  .[, time := .GRP, by = .(date)]
```

```{r netcdf-read-6, cache=TRUE}
# data wrangling
dataPlot = melt(df_summarize, 
                id.vars = c("date", "longitude", "latitude", "time"), 
                variable.name = "model")

# let's comput the ouput of all models
res_anim = ggplot() +
  geom_raster(data = dataPlot,
              aes(x = longitude, y = latitude, fill = value),
              interpolate = TRUE) +
  geom_sf(data = spData::world,
          col = 1,
          fill = "ivory") +
  coord_sf(xlim = c(-165, -115), ylim = c(25, 55)) +
  facet_wrap(. ~ model, ncol = 2) +
  theme_jjo() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = NULL) +
  scale_fill_gradientn(colours = oce::oceColors9A(120)) +
  labs(title = 'Date: {df_summarize[,unique(date)][frame_time]}') +
  transition_time(time) +
  ease_aes('linear')
```
```{r netcdf-read-7, fig.cap="4 models of the current velocity in the North-East part of Pacific ocean"}
# let's print
animate(res_anim, height = 500, width = 500)
```

Four different GIFs, since there are four different models to get the SST:
![[Table from Copernicus](https://catalogue.marine.copernicus.eu/documents/QUID/CMEMS-GLO-QUID-001-031.pdf)](https://i.imgur.com/3OOofoP.png)

```{r netcdf-read-8, eval=FALSE, include=FALSE}
# summarized by week
df_summarize = df[,.(longitude = mean(longitude),
                     latitude = mean(latitude),
                     vel_cglo = mean(vel_cglo)),
                  by = .(date = paste(year(time), '-', week(time)))] %>%
  .[, time := .I]

# let's print the output of CGLO
ggplot() +
  geom_raster(data = df_summarize,
              aes(x = longitude, y = latitude, fill = vel_cglo),
              interpolate = TRUE) +
  geom_sf(data = spData::world,
          col = 1,
          fill = "ivory") +
  coord_sf(xlim = c(-165, -115), ylim = c(25, 55)) +
  theme_jjo()+
  labs(x = NULL, y = NULL) +
  scale_fill_gradientn(colours = oce::oceColors9A(120)) +
  labs(title = 'Date: {df_summarize$date[frame_time]}') +
  gganimate::transition_time(time) +
  gganimate::ease_aes('linear')
```
[^1]: [https://www.unidata.ucar.edu/software/netcdf/](https://www.unidata.ucar.edu/software/netcdf/)
[^2]: [https://www.ogc.org/standards/netcdf](https://www.ogc.org/standards/netcdf)
[^3]: [https://web.itu.edu.tr/~tokerem/netcdf.html](https://web.itu.edu.tr/~tokerem/netcdf.html)
