---
title: "Data Exploration"
author: "Joffrey JOUMAA"
date: "`r format(Sys.Date(), format = '%d %B %Y')`"
output:
  bookdown::html_document2:
    css: cosmo_custom.css
    number_sections: yes
    code_folding: hide
    df_print: default
    fig_caption: yes
    code_download: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
---

```{r setup, include=FALSE}
# global option relative to rmarkdown
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = 'center',
                      message = FALSE)

# library
library(data.table)
library(ggplot2)
library(kableExtra)

# define my own table format: https://github.com/haozhu233/kableExtra/issues/374
sable <- function(x, escape = T, ...) {
  knitr::kable(x, escape = escape, ...) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "responsive"),
      full_width = F
    )
}

# theme ggplot
# based: https://benjaminlouis-stat.fr/en/blog/2020-05-21-astuces-ggplot-rmarkdown/
theme_jjo <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # the whole figure
      #plot.title = element_text(size = rel(1), face = "bold", margin = margin(0,0,5,0), hjust = 0),
      # figure area
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      # axes
      #axis.title = element_text(size = rel(0.85), face = "bold"),
      #axis.text = element_text(size = rel(0.70), face = "bold"),
      axis.line = element_line(color = "black", arrow = arrow(length = unit(0.2, "lines"), type = "closed")),
      # legend
      # legend.title = element_text(size = rel(0.85), face = "bold"),
      # legend.text = element_text(size = rel(0.70), face = "bold"),
      # legend.key = element_rect(fill = "transparent", colour = NA),
      # legend.key.size = unit(1.5, "lines"),
      # legend.background = element_rect(fill = "transparent", colour = NA),
      # Les étiquettes dans le cas d'un facetting
       strip.background = element_rect(fill = "#888888", color = "#888888"),
       strip.text = element_text(size = rel(0.85), face = "bold", color = "white", margin = margin(5,0,5,0))
    )
}
```

This document aims at exploring two datasets, one in 2016 on 6 individuals and another one in 2018 on 4 individuals. For that purpose, we need first to load the `weanlingNES` package to load data.

```{r}
# load library
library(weanlingNES)

# load data
data("data_nes", package = "weanlingNES")
```

# 2016-individuals 

```{r}
# combine all individuals
data_2016 = rbindlist(data_nes$year_2016, use.name = TRUE, idcol = TRUE)
```
 
## Summary

```{r}
# raw_data
data_2016[, .(
  nb_days_recorded = uniqueN(as.Date(date)),
  max_depth = max(maxpress_dbars),
  sst_mean = mean(sst2_c),
  sst_sd = sd(sst2_c)
), by =.id] %>%
  sable(caption="Summary diving information relative to each 2016 individual")
```

Well, it seems that `sst` is a bit odd. Let's have a look at its distribution.


```{r, fig.cap="Distribution of raw `sst2` for the four individuals in 2016"}
ggplot(data_2016, aes(x = sst2_c, fill = .id)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(.id ~ .) +
  theme_jjo()
```

Let's remove any data with a `sst2_c` > 500.

```{r, fig.cap="Distribution of filtered `sst2` for the four individuals in 2016"}
data_2016_filter = data_2016[sst2_c < 500, ]
ggplot(data_2016_filter, aes(x = sst2_c, fill = .id)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(.id ~ .) +
  theme_jjo()
```

Well,that seems to be much better! In the process of filtering out odd values, we removed `r data_2016[,.N] - data_2016_filter[,.N]` rows this way:

```{r}
# nbrow removed
data_2016[sst2_c>500,.(nb_row_removed = .N),by=.id] %>%
  sable(caption = "# of rows removed by indivduals")
```
```{r}
# ggplot(data_2016[,.(.SD, is_out=fifelse(sst2_c>500,T,F))], 
#        aes(x = maxpress_dbars, y=date, col=is_out)) +
#   geom_path(show.legend = FALSE) +
#   facet_wrap(.id ~ .) +
#   theme_jjo()
```

## Some plots

```{r}

```

```{r}
# plot depth records
# ggplot(data_2016, aes(x=date,y=))
```

