---
title: "Changes in Buoyancy Detection"
author: "Joffrey JOUMAA"
date: "`r invisible(Sys.setlocale(locale = 'C')); format(Sys.Date(), format = '%B %d, %Y')`"
output:
  bookdown::html_document2:
    css: cosmo_custom.css
    number_sections: yes
    code_folding: show
    df_print: default
    fig_caption: yes
    code_download: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
vignette: >
  %\VignetteIndexEntry{Changes in Buoyancy Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# global option relative to rmarkdown
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)

# library
library(data.table)
library(ggplot2)
library(magrittr)
library(kableExtra)

# remove some warnings
suppressWarnings(library(ggplot2))

# theme ggplot
# based: https://benjaminlouis-stat.fr/en/blog/2020-05-21-astuces-ggplot-rmarkdown/
theme_jjo <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # the whole figure
      # plot.title = element_text(size = rel(1), face = "bold", margin = margin(0,0,5,0), hjust = 0),
      # figure area
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      # axes
      # axis.title = element_text(size = rel(0.85), face = "bold"),
      # axis.text = element_text(size = rel(0.70), face = "bold"),
      axis.line = element_line(color = "black", arrow = arrow(length = unit(0.2, "lines"), type = "closed")),
      # legend
      # legend.title = element_text(size = rel(0.85), face = "bold"),
      # legend.text = element_text(size = rel(0.70), face = "bold"),
      # legend.key = element_rect(fill = "transparent", colour = NA),
      # legend.key.size = unit(1.5, "lines"),
      # legend.background = element_rect(fill = "transparent", colour = NA),
      # Les <U+00E9>tiquettes dans le cas d'un facetting
      strip.background = element_rect(fill = "#888888", color = "#888888"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "white", margin = margin(5, 0, 5, 0))
    )
}

# define my own table format: https://github.com/haozhu233/kableExtra/issues/374
sable <- function(x, escape = T, ...) {
  knitr::kable(x, escape = escape, ...) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "responsive"),
      full_width = F
    )
}
```

Here, the first thing to do is to retrieve the data processed at the end of `vignette("data_exploration_2018")`.

```{r}
# read the processed data
data_2018_filter <- readRDS("tmp/data_2018_filter.rds")
```

# Daily median drift rate calculation

We then summarize the data by calculating the median of `driftrate` for each day.

```{r}
# calulate the median of driftrate for each day
median_driftrate <- data_2018_filter[divetype == "2: drift",
  .(driftrate = quantile(driftrate, 0.5)),
  by = .(date = as.Date(date), .id)
]

# display 10 random rows
median_driftrate[sample(.N, 10), ] %>%
  sable(caption = "Median of daily drift rate by seals (10 random rows)")
```

```{r fig.cap="Evolution of daily median drift rate across time for each seals"}
# display the result
ggplot(
  median_driftrate,
  aes(x = date, y = driftrate, col = .id)
) +
  geom_point() +
  labs(y = "Daily drift rate (m/s)", x = "Date") +
  theme_jjo() +
  theme(legend.position = "top")
```

# Model daily median drift rate using a [LOESS](https://en.wikipedia.org/wiki/Local_regression)

For each seal, we model the daily median drift rate using a [local polynomial regression](https://en.wikipedia.org/wiki/Local_regression).

> The only parameter to estimate was the span, it was chosen graphically

```{r fig.cap="Evolution of daily median drift rate across time for each seals with a smooth"}
# display the result
ggplot(
  median_driftrate,
  aes(x = date, y = driftrate, col = .id)
) +
  geom_point() +
  geom_smooth(span = 0.25, col = "black") +
  scale_x_date(date_labels = "%m/%Y") +
  labs(y = "Daily drift rate (m/s)", x = "Date") +
  theme_jjo() +
  theme(legend.position = "top") +
  facet_wrap(. ~ .id, scales = "free")
```

# Detection of changes in buoyancy

We finally had to identify when the smooth function change sign.

```{r fig.cap="Evolution of daily median drift rate across time for each seals with a smooth and vertical lines to identify changes in buoyancy"}
# let's identity when the smooth changes sign
changes_driftrate <- median_driftrate %>%
  .[, .(
    y_smooth = predict(loess(driftrate ~ as.numeric(date), span = 0.25)),
    date
  ), by = .id] %>%
  .[c(FALSE, diff(sign(y_smooth)) != 0), ]

# display the result
ggplot(
  median_driftrate,
  aes(x = date, y = driftrate, col = .id)
) +
  geom_point() +
  geom_smooth(span = 0.25) +
  geom_vline(data = changes_driftrate, aes(xintercept = date)) +
  scale_x_date(date_labels = "%m/%Y") +
  labs(y = "Daily drift rate (m/s)", x = "Date") +
  theme_jjo() +
  theme(legend.position = "top") +
  facet_wrap(. ~ .id, scales = "free")
```
